Tidy data structure to support exploration and modeling of temporal-context data

### 2018-11-05 main.Rmd
switched out nesting and crossing

---
title: "Tidy data structure to support exploration and modeling of temporal-context data"
author:
  - familyname: Wang
    othernames: Earo
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : earo.wang@monash.edu
    correspondingauthor: true
  - familyname: Cook
    othernames: Dianne
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : dicook@monash.edu
  - familyname: Hyndman
    othernames: Rob J
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : rob.hyndman@monash.edu
abstract: 'Mining temporal-context data for information is often inhibited by a multitude of time formats: irregular or multiple time intervals, multiple observational units or repeated measurements on multiple individuals, heterogeneous data types, nested and crossed factors indicating hierarchical sub-groups. Time series models, in particular, the software supporting time series forecasting makes strict assumptions on data to be provided, typically a matrix of numeric data with an implicit time index. Going from raw data to model-ready data is painful. This work presents a cohesive and conceptual framework for organizing and manipulating temporal data, which in turn flows into visualization and forecasting routines. Tidy data principles are applied, and extended to temporal data: (1) mapping the semantics of a dataset into its physical layout, (2) including an explicitly declared index variable representing time, (3) incorporating a "key" comprised of single or multiple variables to uniquely identify units over time, using a syntax-based and user-oriented approach in which it imposes nested or crossed structures on the data. This tidy data representation most naturally supports thinking of operations on the data as building blocks, forming part of a "data pipeline" in time-based context. A sound data pipeline facilitates a fluent and transparent workflow for analyzing temporal data. Applications are included to illustrate tidy temporal data structure, data pipeline structure and usage. The infrastructure of tidy temporal data has been implemented in the R package **tsibble**.'
keywords: "temporal data, time series, data structures, data wrangling, tidy data, R, forecasting, data science, exploratory data analysis"
wpnumber: no/yr
jelcodes: C10,C14,C22
blind: false
cover: true
toc: false
bibliography: [references.bib, rpkgs.bib]
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: no
    number_sections: yes
    citation_package: biblatex
---

@@

## Time series of nested and crossed factors

@@

The R package **hts** [@R-hts] is the implementation of structured collections of time series, and provides tools for visualizing, analyzing, and forecasting such data. But it is a pain to configure node list or grouping matrix as the increase in hierarchic depth and grouping factors. This is resulted from the fact that the structure specification is separated from data itself and represented numerically, which is rather counterintuitive from the user's perspective.

@@

The choice of representation of temporal-context data is made from a data-centric perspective, which is taken in the light of the operations that are to be performed on the data. This data abstraction reflects most common problems of incoming data and transformation in reality. Firstly, a data set must be structured in a "tidy" rectangular layout each row has the same schema and each field has a single value. Secondly, declaring the data set to contain temporal observations is determined by an "index" that represents time and a "key" that uniquely identifies each unit that measurements take place on over time. The "key" serves a similar purpose as the panel variable in the Stata's **tsset** command to define the units or subjects, but it is expanded to include multiple variables rather than a single one. A syntax is introduced to express a key consisting of nested and crossed factors. The composition of index and key uniquely defines each observation in a data table, which is equivalent to a primary key [@codd_relational_1970] in a relational database. (The matrix representation also implies that every observation is determined by column variable and time-indexed row.)

Table \@ref(tab:tb-sub) shows a "tidy" temporal data example using a subset of tuberculosis cases estimated by @tb-data. It contains 12 observations and 5 variables arranged in a "long" tabular form. Each observation uniquely records the number of people, who are diagnosed tuberculosis for each gender at three selected countries in the years of 2011 and 2012, where column `year` is declared as the index variable, columns `country`, `continent` and `gender` as key. To be more specific, the key is passed via `id(country | continent, gender)`, which produces a geographic hierarchy of country nested within continent, crossed with a demographic factor. Detailed discussions about index and key specifications will be given in Section \@ref(sec:index) and \@ref(sec:key). Although column `count` is the only measure in this case, it is sufficiently flexible to hold other measured variables, for example, adding the corresponding population size (if known) in order to calibrate the count later.

>>>>

---
title: "Tidy data structure to support exploration and modeling of temporal-context data"
author:
  - familyname: Wang
    othernames: Earo
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : earo.wang@monash.edu
    correspondingauthor: true
  - familyname: Cook
    othernames: Dianne
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : dicook@monash.edu
  - familyname: Hyndman
    othernames: Rob J
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : rob.hyndman@monash.edu
abstract: 'Mining temporal-context data for information is often inhibited by a multitude of time formats: irregular or multiple time intervals, multiple observational units or repeated measurements on multiple individuals, heterogeneous data types. Time series models, in particular, the software supporting time series forecasting makes strict assumptions on data to be provided, typically a matrix of numeric data with an implicit time index. Going from raw data to model-ready data is painful. This work presents a cohesive and conceptual framework for organizing and manipulating temporal data, which in turn flows into visualization and forecasting routines. Tidy data principles are applied, and extended to temporal data: (1) mapping the semantics of a dataset into its physical layout, (2) including an explicitly declared index variable representing time, (3) incorporating a "key" comprised of single or multiple variables to uniquely identify units over time. This tidy data representation most naturally supports thinking of operations on the data as building blocks, forming part of a "data pipeline" in time-based context. A sound data pipeline facilitates a fluent and transparent workflow for analyzing temporal data. Applications are included to illustrate tidy temporal data structure, data pipeline structure and usage. The infrastructure of tidy temporal data has been implemented in the R package **tsibble**.'
keywords: "temporal data, time series, data structures, data wrangling, tidy data, R, forecasting, data science, exploratory data analysis"
wpnumber: no/yr
jelcodes: C10,C14,C22
blind: false
cover: true
toc: false
bibliography: [references.bib, rpkgs.bib]
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: no
    number_sections: yes
    citation_package: biblatex
---

@@

<!--
## Time series of nested and crossed factors

@@

The R package **hts** [@R-hts] is the implementation of structured collections of time series, and provides tools for visualizing, analyzing, and forecasting such data. But it is a pain to configure node list or grouping matrix as the increase in hierarchic depth and grouping factors. This is resulted from the fact that the structure specification is separated from data itself and represented numerically, which is rather counterintuitive from the user's perspective.
-->

@@

The choice of representation of temporal-context data is made from a data-centric perspective, which is taken in the light of the operations that are to be performed on the data. This data abstraction reflects most common problems of incoming data and transformation in reality. Firstly, a data set must be structured in a "tidy" rectangular layout each row has the same schema and each field has a single value. Secondly, declaring the data set to contain temporal observations is determined by an "index" that represents time and a "key" that uniquely identifies each unit that measurements take place on over time. The "key" serves a similar purpose as the panel variable in the Stata's **tsset** command to define the units or subjects, but it is expanded to include multiple variables rather than a single one. The composition of index and key uniquely defines each observation in a data table, which is equivalent to a primary key [@codd_relational_1970] in a relational database. (The matrix representation also implies that every observation is determined by column variable and time-indexed row.)

Table \@ref(tab:tb-sub) shows a "tidy" temporal data example using a subset of tuberculosis cases estimated by @tb-data. It contains 12 observations and 5 variables arranged in a "long" tabular form. Each observation uniquely records the number of people, who are diagnosed tuberculosis for each gender at three selected countries in the years of 2011 and 2012, where column `year` is declared as the index variable, columns `country`, `continent` and `gender` as key. Detailed discussions about index and key specifications will be given in Section \@ref(sec:index) and \@ref(sec:key). Although column `count` is the only measure in this case, it is sufficiently flexible to hold other measured variables, for example, adding the corresponding population size (if known) in order to calibrate the count later.
